---
title: "Reversing the Stroop Effect"
author: "220225579"
date: "`r Sys.Date()`"
output: html_document
---

## Project Motivation  

My aim with the project was to attempt communicating psychological theory in an accessible way. The Stroop Effect was the first lab experiment I ever did in my undergrad Psych course, and it blew my mind. Secondly, having worked with students, it has been my mission to spread the gospel of distributed practice. I hope this assessment will reach more young people (and adults*) and convince them of the wonders of regular learning sessions.

*me

## Data Origins  

The data was sourced from *Cognitive Dynamics of a Single Subject: 1428 Stroop Tests and Other Measures in a Mindfulness Meditation Context Over 2.5 Years* by Matti Toivo Juhani Heino (2022), [DOI: 10.5334/jopd.51](https://openpsychologydata.metajnl.com/articles/10.5334/jopd.51).

God bless this person's patience!

## Psych 101

### What is the Stroop Effect?  

Try saying the *colour* the words are written in (not the word):

**$\color{red}{\text{red}}$**  

**$\color{blue}{\text{blue}}$**  

**$\color{green}{\text{green}}$**  

**$\color{purple}{\text{purple}}$**     



That was simple enough.  



Now try that again: say the font colour, not the word:

**$\color{green}{\text{red}}$**   

**$\color{red}{\text{blue}}$**    

**$\color{purple}{\text{green}}$**    

**$\color{blue}{\text{purple}}$**   


Did it take longer this time? That's because of the *Stroop Effect*. It's named after the John Stroop, who realised that it's faster to say the colour names when the text matches the colour [(Stroop, 1935)](https://psych.hanover.edu/classes/Cognition/papers/stroop%201933.pdf).  

But what if there was a way to get better at the Stroop Task?

That brings us to...   

### What is distributed practice?  

Distributed practice involves short and frequent study sessions over a longer period of time, and it has been shown to improve recalling information [(Cepeda et al, 2006)](http://www.yorku.ca/ncepeda/publications/CPVWR2006.pdf).  

This is why teachers don't like last-minute cram sessions - it doesn't work in the long term!    

### Research Questions  

Can spaced practice improve performance on the Stroop task?

### Preparing the data

```{r error=FALSE, message=FALSE}
#setting the working directory
setwd("C:/Users/piyal/OneDrive/Desktop/R/psy6422/PSY6422")

#load the relevant libraries
library(tidyverse)
library(here)
library(ggplot2)
```

```{r}
#load the data
source <- here("data", "cognitive_dynamics_heino.csv")
stroop <- read.csv(source)
```

There are too many variables in the original dataset, so I would like choose the ones relevant for the visualisation.  

And also change the names of the columns to make my coding life easier.  

```{r}
#changing column names to make them user/coder friendly
stroop_clean <- stroop %>%
  select(date, stroop_congruent_pre_meditation, stroop_incongruent_pre_meditation) %>%
   rename(congruent = stroop_congruent_pre_meditation,
         incongruent = stroop_incongruent_pre_meditation)

#date is a character, needs to be made into date format
stroop_clean <- stroop_clean %>% 
  mutate(date = dmy(date)) %>%                 #changing string to date
  filter(date > "2014-07-01", na.rm = TRUE)    #deleting dates that are outliers


#changing centiseconds to milliseconds
stroop_clean <- stroop_clean %>% 
  mutate(congruent = congruent * 10,
         incongruent = incongruent * 10)

#reshaping dataframe to long
stroop_long <- stroop_clean %>% select(date, congruent, incongruent) %>% 
  gather(condition, time, congruent, incongruent)
```

Let's have a look at our final dataset!

```{r}
head(stroop_long)
```

That looks neat and concise and ready for the next step!

### Visualising the data

I wanted to make a neater visualisation by making some adjustments to the basic ggplot theme. 

```{r error=FALSE, message=FALSE}
#defining the theme
theme_stroop <- function(){
 
  theme_minimal() %+replace%         #replacing the elements I want to change
  
  theme(
    
    #grid elements
    panel.grid.minor = element_blank(),
    
    #text elements
    plot.title = element_text(      #title
      size = 15,                    #set font size
      face = 'bold',                #bold typeface
      hjust = 0,                    #left align
      vjust = 2),                   #raise slightly

    plot.subtitle = element_text(   #subtitle
      size = 10,                    #font size
      hjust = 0),                   #left align
    
    axis.title = element_text(      #axis titles
      size = 8),                    #font size
    
    axis.text = element_text(       #axis text
      size = 8),                    #font size
    
    plot.caption = element_text(    #caption
      size = 8,                     #font size
      hjust = 1),                   #right align
    
    legend.position = c(.9,.8),  #positioning legend
    
    plot.background = element_rect(fill = "white", color = NA)
    )
}

#building a colour palette based on the original Stroop colours

palette_stroop <- c("#ff0000", "#0085ff", "#3cff00", "#8c00db", "#543324")

```


And now, the moment we have been waiting for!


```{r eval=TRUE, error=FALSE, message=FALSE, out.width="100%"}
#plotting reaction time for congruent and incongruent trials
p1 <- ggplot(data = stroop_long,
             mapping = aes(x = date,
                           y = time,
                           colour = condition))

plot1 <- p1 + 
  geom_smooth(method = "lm", se = FALSE, na.rm = TRUE)+                # linear regression lines
  geom_point(alpha = 0.5, na.rm = TRUE) +                              #scatter plot
  theme_stroop() +                                                     #the theme defined previously
  labs(x = "Year",                                                     #labelling the graph
       y = "Reaction Time in ms",
       title = "Distributed Practice Improves Performance on Stroop Task",
       subtitle = "Time across 2.5 years vs time taken for each Stroop trial",
       caption = "Source: 'Cognitive dynamics of a single subject'\n Matti Toivo Juhani Heino")+
  scale_colour_manual(values = palette_stroop) +                       #using the colours defined previously
  scale_x_date(date_breaks = "4 months", date_labels = "%b %Y")        #defining the intervals on the x-axis

plot1                                                                  #let's take a look at our plot
```

it looks like practice does make you faster at both tasks - reading the congruent and incongruent words!  

Wait, there's more! Does deliberate practice actually **reverse the Stroop Effect?!**  

Let's take a closer look.  

```{r error=FALSE, message=FALSE, out.width="100%"}
#creating a magnified inset showing crossover

#coordinates of inset
x0 <- as.Date("2016-01-01", "%Y-%m-%d")
x1 <- as.Date("2016-09-01", "%Y-%m-%d")
y0 <- 140
y1 <- 180

#coordinates of origin
x3 <- as.Date("2016-04-01", "%Y-%m-%d")
x4 <- as.Date("2016-06-30", "%Y-%m-%d")
y3 <- 90
y4 <- 100

#coordinates of section to subset
x01 <- as.Date("2016-04-01", "%Y-%m-%d")
x02 <- as.Date("2016-06-15", "%Y-%m-%d")
  
y01 <- 70
y02 <- 140
  
#the inset done manually
p2 <- ggplot(data = stroop_long, aes(x = date,                    #basic plot
                                     y = time,
                                     colour = condition)) +
  geom_smooth(method = "lm", se = FALSE, na.rm = TRUE) +          #regression line
  xlim(x01, x02) +                                                #subsetting the part that crosses over
  theme_stroop()+                                                 #use our theme from before
  scale_colour_manual(values = palette_stroop)+                   #using the colours from before
  theme(legend.position = "none") +                               #removing the legend in the inset
  theme(panel.background = element_rect(fill = "ghostwhite"),     #giving it a background to highlight it
        axis.text.y = element_blank(),                            #removing the y-axis text
        axis.title = element_blank())                             #removing the titles from the axes


#adding the inset to the original plot  
plot2 <- plot1 + 
  annotation_custom(ggplotGrob(p2),    #adding the inset
                    xmin = x0,         #defining the coordinates for the inset
                    xmax = x1,
                    ymin = y0,
                    ymax = y1) +
  
  geom_rect(aes(xmin = x0,              #border of inset
                xmax = x1,
                ymin = y0,
                ymax = y1),
              colour = "black",
           linewidth = .6,
               alpha = 0) +
  
  geom_rect(aes(xmin = x3,              #border highlighting the section on graph
                xmax = x4,
                ymin = y3,
                ymax = y4),
              colour = "black",
           linewidth = .6,
                alpha = 0) +
  
     annotate("segment", x = x3, xend = x0,   #connecting graph to inset
                         y = y4, yend = y0,   
             colour = "black",
             linewidth = .6,
             linetype = 2) +
    annotate("segment", x = x4, xend = x1,    #connecting graph to inset
                        y = y4, yend = y0,
           colour = "black",
           linewidth = .6,
           linetype = 2) 

plot2
  
```

```{r}
#saving our plot
ggsave(here("figs", "stroop_practice_inset.png"), 
       plot = plot2, width = 25, height = 15, units = "cm")
```

This graph shows the effect of daily practice on the Stroop Effect.   

With enough practice you could end up reading this:

**$\color{green}{\text{red}}$**   

**$\color{red}{\text{blue}}$**    

**$\color{purple}{\text{green}}$**    

**$\color{blue}{\text{purple}}$**  

...faster than you read this:  

**$\color{red}{\text{red}}$**  

**$\color{blue}{\text{blue}}$**  

**$\color{green}{\text{green}}$**  

**$\color{purple}{\text{purple}}$**    

Imagine what else you could become proficient in with distributed practice!

### Further explorations

The original dataset also has data on meditation, the time of day when the Stroop task was done. It would be interesting to see how those factors play a role in learning.